---
title: "HW_8"
author: "Greg_G"
date: "2022-08-09"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Import dplyr and read in data:

```{r pressure, echo=FALSE}
library(dplyr)

df <- read.csv("hw8_data.csv", check.names = FALSE)
```

Questions:
1. How many tissue samples are there by donor and what is their disease status?
  • Display your answer as a data frame.
  • Disease status should be ordered as follows: Normal, Potential, Affected.

```{r}
df_counts <- df %>%
  group_by(donor) %>%
  summarize(`Number of Tissues` = n_distinct(Tissue),
            `Disease Status` = unique(`Disease Status`)) %>%
  arrange(factor(`Disease Status`, levels = c("Normal", "Potential", "Affected")))

df_counts
```

  
2. What is the average AND standard deviation values for CD45+ and PBMCs (%CD45) when grouped by disease status and tissue?
  • Display your answer as a data frame.
  • Disease status should be ordered as follows: Normal, Potential, Affected.
```{r}
df_CD45 <- df %>%
  group_by(`Disease Status`, Tissue) %>%
  summarize("Mean CD45+" = mean(`CD45+`),
            "SD CD45+" = sd(`CD45+`),
            "Mean PBMCs (%CD45)" = mean(`CD45+`),
            "SD PBMCs (%CD45)" = sd(`PBMCs (%CD45)`))

df_CD45
```
  
3. What is the mean difference between the Eosinophil populations between the different disease statuses, irrespective of tissue?
  • Display your answer as a data frame.
  • Disease status should be ordered as follows: Normal, Potential, Affected.
```{r}
df_Eos <- df %>%
  group_by(`Disease Status`) %>%
  summarize("Mean Eosinophils" = mean(Eosinophils)) %>%
  arrange(factor(`Disease Status`, levels = c("Normal", "Potential", "Affected")))

df_Eos <- df_Eos %>%
  mutate("Mean Diff Normal" = `Mean Eosinophils` - `Mean Eosinophils`[1]) %>%
  mutate("Mean Diff Potential" = `Mean Eosinophils` - `Mean Eosinophils`[2]) %>%
  mutate("Mean Diff Affected" = `Mean Eosinophils` - `Mean Eosinophils`[3])

df_Eos
```

4. When stratified by tissue and disease status, which donors have the highest CD8 Tem CD69+ proportions?
  • Display your answer as a data frame.
  • Disease status should be ordered as follows: Normal, Potential, Affected.
```{r}
df_Tem <- df %>%
  group_by(`Disease Status`, Tissue) %>%
  summarize("Donor Max CD8 Tem CD69+" = donor[which(`CD8 Tem CD69+` == max(`CD8 Tem CD69+`))]) %>%
  arrange(factor(`Disease Status`, levels = c("Normal", "Potential", "Affected")))

df_Tem
```

5. What are the average AND standard deviation values for all columns that start with “CD4 Tem” when grouped by disease status and tissue?
  • Display your answer as a data frame.
  • Disease status should be ordered as follows: Normal, Potential, Affected.
  • There are 26 columns that start with “CD4 Tem”. You should not be manually specifying
  “mean(‘CD4 Tem blah blah‘)” for each of those 26 columns. dplyr has additional ways to solve this.
```{r}
df_CD4Tem <- df %>%
  group_by(`Disease Status`, Tissue) %>%
  summarise(across(starts_with("CD4 Tem"), c(mean, sd))) %>%
  rename_with(.cols = ends_with("_1"), function(x) paste0(x, " Mean")) %>%
  rename_with(.cols = ends_with("_2"), function(x) paste0(x, " SD")) %>%
  rename_with(function(x) gsub("_.", "", x)) %>%
  arrange(factor(`Disease Status`, levels = c("Normal", "Potential", "Affected")))

df_CD4Tem
```
